---
title: "354/454-Final-Exam"
subtitle: "YOUR NAME"
format:
  revealjs: 
    theme: slides.scss
    smaller: true
    scrollable: true
    slide-number: true
    chalkboard:
      buttons: true
    preview-links: auto
    footer: YOUR NAME OR TOPIC OR DELETE
resources:
  - demo.pdf
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  cache = FALSE, # if TRUE knitr will cache results to reuse in future knits
  fig.width = 6, # the width for plots created by code chunk
  fig.height = 4.5, # the height for plots created by code chunk
  fig.align = 'center', # how to align graphics. 'left', 'right', 'center'
  dpi = 300,
  dev = 'png',
  # results = 'asis', # knitr passes through results without reformatting
  echo = TRUE, # if FALSE knitr won't display code in chunk above it's results
  message = TRUE, # if FALSE knitr won't display messages generated by code
  strip.white = TRUE, # if FALSE knitr won't remove white spaces at beg or end
  warning = FALSE) # if FALSE knitr won't display warning messages in the doc

```

# Include your style guide here {-}

```{r, styleguide}


```

# Final Exam - 25 Points Total

## Please choose **one** of the following datasets for the exam:

- [Tidy Tuesday - LEGO Database](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-09-06)
- [Tidy Tuesday - Passwords](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-01-14/readme.md)
- [Tidy Tuesday - Stanford Open Policing Project](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-03-19)
- Or Any other Tidy Tuesday Dataset

## Additional Information

This does not mean that you are entirely limited to these datasets, if you wish to find more data related to one of the sets then you may use it **IN ADDITION** to what is provided. Note that finding extra data **is not required**.

You will have until the day of the final exam to become familiar with your dataset. By the final exam slot, you should have created 3 (or more) well-thought, detailed, and visually appealing graphics to present to the class. Please create a polished .qmd presentation that will **last for 15 minutes.** All students will present during the 3 hours for the exam with grades being based on peer and professor/TA opinions.

## Talk about your dataset

Please put effort into these and **know** your dataset. These are not just visuals to throw up, you are there to teach people and tell a story.

You are telling a story about your dataset.

## Who is your audience

Please Tell us who you believe your audience is before your presentation.

- general audience
- scientific audience
- data enabled audience
- etc.